---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nicher

<!-- badges: start -->
<!-- badges: end -->

The goal of nicher is to create ecological niche models based on
an ellipse model

## Installation

You can install the development version of nicher from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("alrobles/nicher")
```

## Example

This is a basic example which shows you how to solve a common problem. First we
get the shapefile of an accesibility area (M):

```{r example}
library(nicher)
library(terra)


M_path <- system.file("extdata", "Mshp_test.rds", package="nicher")
Mshp <- terra::unwrap(readr::read_rds(M_path))
plot(Mshp)
```


Then we get environmental variables to model:
```{r}

stack_path <- system.file("extdata", "stack_1_12_crop.rds", package="nicher")
# 2 variables
stack_1_12 <- terra::unwrap(readr::read_rds(stack_path))
plot(stack_1_12)
# 3 variables
stack_1_12_19 <- get_example_data("stack_1_12_19")
plot(stack_1_12_19)
```

We get the parameters of the ellipse:

```{r}
pars_2var <-  get_ENM_par(rawSpOccPnts, stack_1_12, Mshp, method = "mahalanobis")

pars_3var <-  get_ENM_par(rawSpOccPnts, stack_1_12_19, Mshp, method = "mahalanobis")
```


Then we plot the ellipse (2 vars case):
```{r 2varsPars,  echo = FALSE}
plot(pars_2var)
```

Then we plot the ellipse (3 vars case):
```{r 3varsPars,  echo = FALSE}
plot(pars_3var)
```

We predict the suitability given environmental data (2 vars case):
```{r predict2vars, echo = FALSE}
outputRast_2vars <- predict(pars_2var, stack_1_12)
plot(outputRast_2vars)

```

Finally we predict the suitability given environmental data (3 vars case All world):
```{r predict3vars, echo = FALSE}
outputRast_3vars <- predict(pars_3var, stack_1_12_19)
plot(outputRast_3vars)

```

