---
title: "Cross-validation for penalised niche models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-validation for penalised niche models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(nicher)
```

## Why this vignette exists

`nicher 3.0`–`3.2` give you seven likelihood families
(`presence_only`, `weighted`, `kde_bias_corrected`, `skew_normal`,
`skew_normal_weighted`, `skew_t`, `skew_t_weighted`) and three penalty
knobs (`prior_mu_lambda`, `prior_log_sigma_lambda`,
`prior_alpha_lambda`). That gives you a lot of dials. This vignette
explains:

1.  Why the unregularised `weighted` MLE is **pathological** on real
    data, using the bundled `example_vicugna` as the case study.
2.  Why ordinary AIC / BIC (added in `nicher 3.2`) **cannot** tell you
    how strong a penalty to use.
3.  How `cv_nicher()` fixes that and lets you tune the penalty
    knobs principled-ly.
4.  How to combine `cv_nicher()` with `compare_nicher()` so you can
    pick both the best **family** and the best **penalty strength**.

We use frequentist tooling throughout. There is no MCMC, no posterior
sampling, no marginal likelihood. Everything below is penalised
maximum likelihood with cross-validated tuning.

## 1. The pathology: unpenalised `weighted` on Vicugna

The Vicugna dataset (`example_vicugna`) has 545 occurrences of
*Vicugna vicugna* in two bioclimatic variables (`bio1`, `bio12`) and
10 000 background points sampled from the species' accessible region
(*M*).

```{r vicugna-data}
library(nicher)
env_occ <- example_vicugna$env_occ
env_m   <- example_vicugna$env_m
```

Fit a **presence-only** baseline (no penalty needed, no `env_m`):

```{r vicugna-po}
fit_po <- optimize_niche(
  env_occ = env_occ, env_m = NULL,
  num_starts = 30L, breadth = 0.1,
  likelihood = "presence_only", verbose = FALSE
)
mu_po    <- fit_po$best$theta[1:2]
sigma_po <- exp(fit_po$best$theta[3:4])
mu_po
#> [1] 6.118 214.958
sigma_po
#> [1]  3.019 233.971
```

Now the **unpenalised** weighted fit (the "paper" version of Jiménez
et al. 2022 with all penalty knobs set to zero):

```{r vicugna-weighted-unpen}
fit_w_unpen <- optimize_niche(
  env_occ = env_occ, env_m = env_m,
  num_starts = 30L, breadth = 0.45,
  likelihood = "weighted",
  prior_mu_lambda        = 0,        # all penalties OFF
  prior_log_sigma_lambda = 0,
  prior_alpha_lambda     = 0,
  verbose = FALSE
)
fit_w_unpen$best$theta
#>        mu1        mu2 log_sigma1 log_sigma2     s_par1
#> -47.984    679.623     3.032      6.120     -0.737
```

The recovered niche centre is `mu = (-48, 680)` while the data has
mean `(6.1, 215)`. This is **wildly outside the data cloud**.
`log σ_2 ≈ 6.1` corresponds to `σ_2 ≈ 451`, much larger than the data
SD of 234. The fit has drifted out of the support of the data.

### What happened?

The "weighted" likelihood (Eq. 5 of Jiménez et al. 2022) is

\[
\ell(\theta) \;=\; \sum_{i=1}^{n_{\rm occ}} \log f(x_i;\theta)
              \;-\; n_{\rm occ}\,\log
              \left[ \sum_{j=1}^{n_M} w(y_j)\,f(y_j;\theta) \right]
\]

where $f$ is the parametric niche density (Gaussian, skew-normal, …)
and the denominator integrates the niche over the accessible region
*M*.

This likelihood has a **well-known unbounded-MLE pathology** when *M*
covers a much wider region than the occurrences (Patil & Ord 1976;
Jiménez et al. 2022 §3.4 acknowledge this without solving it):

*   You can make the **denominator** $\sum_j w(y_j) f(y_j;\theta)$
    arbitrarily small by inflating $\Sigma$ — a flat enough bell curve
    over the M cloud places little mass on any single point.
*   You can also make it small by **moving μ outside the M cloud**
    altogether (the curve becomes mostly tails over M).
*   Meanwhile the **numerator** $\sum_i \log f(x_i;\theta)$ only
    decays quadratically as μ drifts away from the occurrences and
    $\Sigma$ inflates.

The ratio numerator / denominator can therefore go to **+∞** along a
trajectory where μ leaves the data and $\Sigma$ inflates. The
optimiser finds that ridge in the likelihood surface and walks off.
What you see is exactly that: μ at $(-48, 680)$ with σ inflated.

This is **not a bug**. The likelihood is genuinely unbounded above on
finite data; the MLE does not exist. The fix is regularisation.

### Quick sanity check

Compute the un-penalised log-likelihood at both fits:

```{r vicugna-loglik-comparison}
logLik(fit_po)           # presence-only
logLik(fit_w_unpen)      # weighted, no penalty
```

The un-penalised weighted loglik will be **larger** (less negative)
than the PO loglik. That's the definition of the pathology: the
"better" fit is the one that has drifted out of the data.

## 2. Why AIC and BIC cannot tune the penalty

`nicher 3.2` provides `AIC()`, `BIC()`, and `compare_nicher()`. They
are perfect for ranking different **likelihood families** at fixed
penalty knobs. They are **useless** for tuning the penalty knobs
themselves.

The reason is mechanical: $\text{AIC} = -2\,\ell(\hat\theta) + 2k$
where $k = \text{length}(\hat\theta)$ is constant across penalty
strengths (same family ⇒ same $k$). And the unpenalised loglik
$\ell(\hat\theta_\lambda)$ at the converged $\hat\theta_\lambda$ is
**monotone-decreasing in $\lambda$** — any penalty pulls
$\hat\theta$ away from the unpenalised MLE, which by definition has
the largest unpenalised loglik. Combine: AIC is **monotone-increasing
in $\lambda$**, so the AIC-best model is always $\lambda = 0$, which
is the pathological fit.

You can verify this directly. Fit `weighted` at $\lambda_\mu \in \{0,
1, 10\}$:

```{r aic-fails}
fit_lambda0  <- fit_w_unpen
fit_lambda1  <- optimize_niche(
  env_occ, env_m, num_starts = 30L, breadth = 0.45,
  likelihood = "weighted",
  prior_mu_lambda = 1, prior_log_sigma_lambda = 1, prior_alpha_lambda = 0,
  verbose = FALSE
)
fit_lambda10 <- optimize_niche(
  env_occ, env_m, num_starts = 30L, breadth = 0.45,
  likelihood = "weighted",
  prior_mu_lambda = 10, prior_log_sigma_lambda = 10, prior_alpha_lambda = 0,
  verbose = FALSE
)
compare_nicher(
  l0  = fit_lambda0,
  l1  = fit_lambda1,
  l10 = fit_lambda10
)
#>   model likelihood     loglik df nobs    AIC   dAIC ...
#> 1    l0   weighted   <best>    5  545   <min>   0
#> 2    l1   weighted   <worse>   5  545   ...     >0
#> 3   l10   weighted   <worst>   5  545   ...     >0
```

The pathological $\lambda = 0$ fit "wins" by AIC. We need a different
metric.

There **is** a workaround if you really want IC-style numbers: use
`compare_nicher(..., comparison_basis = "penalised")`, which computes
AIC from the **penalised** objective $\ell(\hat\theta) -
\text{pen}(\hat\theta)$. That score is no longer monotone in
$\lambda$, but it conflates "model fit" with "penalty preference",
which is hard to interpret. For tuning $\lambda$, cross-validation is
the cleaner tool.

## 3. Cross-validation breaks the deadlock

The principled tool is **out-of-sample log-likelihood**. Hold back a
fold of `env_occ`, refit on the remainder, score the fold using the
fold's converged $\hat\theta_\lambda$. The pathological fit $\hat\theta_0$
puts μ at $(-48, 680)$, which is far from any occurrence; held-out
occurrences therefore have **catastrophic** density. A
moderately-regularised fit stays close to the data cloud and
therefore predicts held-out occurrences well.

`cv_nicher()` automates this:

```{r cv-vicugna}
cv0  <- cv_nicher(fit_lambda0,  env_occ, env_m, type = "kfold", k = 5L, seed = 1L)
cv1  <- cv_nicher(fit_lambda1,  env_occ, env_m, type = "kfold", k = 5L, seed = 1L)
cv10 <- cv_nicher(fit_lambda10, env_occ, env_m, type = "kfold", k = 5L, seed = 1L)

c(lambda_0  = cv0$cv_loglik,
  lambda_1  = cv1$cv_loglik,
  lambda_10 = cv10$cv_loglik)
#> lambda_0  lambda_1  lambda_10
#>   <very neg>   <best>    <middle>
```

The CV log-likelihood is **U-shaped in $\lambda$**:

*   At $\lambda = 0$ the fit overfits / drifts out of data ⇒ held-out
    points have near-zero density ⇒ very bad CV loglik.
*   At $\lambda \to \infty$ the fit is pinned to the PO baseline,
    ignoring `env_m` entirely ⇒ moderate CV loglik (no worse than
    pure PO, no better either).
*   At intermediate $\lambda$ the fit uses `env_m` to correct PO
    without drifting away ⇒ best CV loglik.

This is the curve you sweep to **set the penalty strength** in a
principled way. The minimum-CV-error $\lambda$ is your selected
regularisation.

### A small grid search

```{r cv-grid}
lambdas <- c(0, 0.1, 0.3, 1, 3, 10, 30)
fits <- lapply(lambdas, function(lam) {
  optimize_niche(
    env_occ, env_m, num_starts = 30L, breadth = 0.45,
    likelihood = "weighted",
    prior_mu_lambda = lam, prior_log_sigma_lambda = lam,
    verbose = FALSE
  )
})
cvs <- vapply(fits, function(f) {
  cv_nicher(f, env_occ, env_m, type = "kfold", k = 5L, seed = 1L)$cv_loglik
}, numeric(1L))
data.frame(lambda = lambdas, cv_loglik = cvs)
```

The row with the largest `cv_loglik` is your tuned model. Use that
$\lambda$ for downstream prediction (`predict.nicher()`,
`habitat_suitability()`), reporting, and any further analysis.

## 4. Combining `cv_nicher` with `compare_nicher`

Workflow for a typical study:

1.  **For each likelihood family** (`presence_only`, `weighted`,
    `skew_normal_weighted`, `skew_t_weighted`, …) run a small CV grid
    over the penalty knobs to find the family's best $\lambda^\star$.
2.  Fit each family at its $\lambda^\star$.
3.  `compare_nicher()` those fits (with default
    `comparison_basis = "unpenalised"`) to pick the best family.

The two stages are doing different jobs. Stage 1 (CV) tunes
**continuous regularisation parameters within a family**. Stage 2
(AIC/BIC) ranks **discrete model structures across families**. AIC and
BIC handle stage 2 well precisely because the families have different
$k$ — Akaike's `+2k` correction is doing real work. But for stage 1
(fixed family, varying $\lambda$, fixed $k$) the `+2k` correction is
useless and CV is the only sound tool.

## 5. Practical notes

### Choosing the fold scheme

`type = "kfold"` with `k = 5` (default) is the standard choice. For
small `env_occ` (`n < 100`) consider `k = 10` or `type = "loo"` to
reduce fold-noise. LOO is exhaustive and unbiased but expensive: `n`
refits per evaluation. Vicugna's 545 occurrences with `num_starts_cv =
1L` (warm-started from the full-data $\hat\theta$) takes about
30 seconds for `weighted` and 5 minutes for `skew_t_weighted`.

### When to bump `num_starts_cv`

`cv_nicher()` defaults to `num_starts_cv = 1L`, warm-starting each
fold from the full-data $\hat\theta$. This is fast but assumes the
per-fold likelihood is unimodal **near** $\hat\theta$. For dense data
this almost always holds. For sparse data, or for the skew families
where the surface can have multiple basins, set `num_starts_cv = 5L`
or more — the per-fold optimization will then use the same Sobol
multistart machinery as the original `optimize_niche()` call, at the
cost of `num_starts_cv ×` more wall time per fold.

### Spatial autocorrelation

Random k-fold CV assumes the occurrences are i.i.d. exchangeable. For
spatially clustered data the held-out fold often has near-neighbours
in the train fold, leading to **optimistic** CV loglik (the model is
easy to predict). The standard remedy is **spatial-block CV**: assign
folds based on spatial location, so each held-out block is far from
its train data. `cv_nicher()` does not yet support spatial blocks; it
is on the roadmap. For now, treat random-CV scores as upper bounds on
generalisation performance, especially for highly clustered datasets.

### Effective degrees of freedom

A more theoretical alternative to CV is **AIC with effective degrees
of freedom**: $\text{AIC}_{\rm eff} = -2\ell + 2 \cdot
\text{tr}(H^{-1}_{\rm pen} H_{\rm data})$. The trace term is the
"effective" number of parameters after the ridge has shrunk some of
them. As $\lambda \to \infty$, $\text{tr} \to 0$; at $\lambda = 0$,
$\text{tr} = k$. This makes AIC non-monotone in $\lambda$ in the
right way. It is on the roadmap (it requires the data Hessian, which
the package can compute via finite differences), but for now CV is
the only built-in option for tuning $\lambda$.

### Forward compatibility with nicherbayes

The CV machinery here computes
$\sum_{\rm folds} \log p(x_{\rm test} \mid \hat\theta_{\rm train})$.
A future Bayesian sibling package would compute the **posterior
predictive** $\sum_{\rm folds} \log \int p(x_{\rm test} \mid \theta)\,
p(\theta \mid x_{\rm train})\,d\theta$ via importance sampling or
similar. The fold structure, the fingerprint validation, and the
warm-start logic are all reusable. This vignette's CV is a frequentist
approximation to that procedure; the difference vanishes under a flat
prior on $\theta$ around $\hat\theta_{\rm train}$.

## References

*   Akaike, H. (1973). Information theory and an extension of the
    maximum likelihood principle. *2nd International Symposium on
    Information Theory*.
*   Hastie, T., Tibshirani, R. (1990). *Generalized Additive Models*,
    Chapman & Hall.
*   Jiménez, L., Soberón, J., Christen, J.A., Soto, D. (2022). On
    the problem of modeling a fundamental niche from occurrence data.
    *Ecological Modelling* 414, 108890.
*   Patil, G.P., Ord, J.K. (1976). On size-biased sampling and
    related form-invariant weighted distributions. *Sankhyā* 38B,
    48-61.
*   Stone, M. (1974). Cross-validatory choice and assessment of
    statistical predictions. *J. Royal Statistical Society* B 36,
    111-147.
*   Wood, S.N. (2017). *Generalized Additive Models: An Introduction
    with R*, 2nd ed., Chapman & Hall.
