---
title: "Package Workflow: End-to-End Niche Fitting and Projection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Workflow: End-to-End Niche Fitting and Projection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette walks through the complete `nicher` workflow: loading
environmental data, fitting niche models, comparing them with information
criteria, projecting habitat suitability onto geographic space, and
extracting the estimated ellipsoid parameters.

## 1 Load data

The package ships with cropped rasters and occurrence records for the
Emerald-bellied Hummingbird (*Abeillia abeillei*).

```{r load-data}
library(nicher)
library(terra)

stack_path <- system.file("extdata", "stack_1_12_crop.rds",
                          package = "nicher")
env_rasters <- readRDS(stack_path)
data(example_occ_df, package = "nicher")
```

## 2 Extract environmental values

Occurrence records are longitude/latitude coordinates. We extract raster
values at those points (`env_occ`) and use every non-`NA` cell as the
accessible background (`env_m`).

```{r extract}
occ_points <- terra::vect(example_occ_df, geom = c("lon", "lat"),
                          crs = "EPSG:4326")

env_occ <- terra::extract(env_rasters, occ_points, ID = FALSE)
env_occ <- stats::na.omit(as.data.frame(env_occ))

env_m <- stats::na.omit(as.data.frame(terra::values(env_rasters)))

str(env_occ)
str(env_m)
```

## 3 Fit niche models

`optimize_niche()` is the main entry point. We fit a presence-only model
and an inverse-weighted model that corrects for sampling bias in the
accessible environment.

```{r fit, results = "hold"}
set.seed(42)

fit_po <- optimize_niche(
  env_occ    = env_occ,
  env_m      = NULL,
  likelihood = "presence_only",
  num_starts = 3L,
  seed       = 42L,
  control    = list(maxeval = 1000L),
  verbose    = FALSE
)

fit_ipw <- optimize_niche(
  env_occ    = env_occ,
  env_m      = env_m,
  likelihood = "ip_weighted",
  num_starts = 3L,
  seed       = 42L,
  control    = list(maxeval = 150L),
  verbose    = FALSE,
  grad       = "analytic"
)

fit_po
fit_ipw
```

## 4 Convergence diagnostics

The `assess()` generic checks whether multiple starts converged to the
same optimum. A `"converged"` flag means all starts agree within
tolerance.

```{r assess}
assess(fit_po)$flag
assess(fit_ipw)$flag
```

## 5 Model comparison

`compare_nicher()` tabulates log-likelihood, AIC, BIC, and Akaike
weights across competing fits.

```{r compare, warning = FALSE}
cmp <- compare_nicher(
  presence_only  = fit_po,
  ip_weighted    = fit_ipw
)
print(cmp[, c("model", "likelihood", "AIC", "dAIC", "BIC", "dBIC")])
```

## 6 Extract ellipsoid parameters

`get_ellipsoid_pars()` computes the sample mean and covariance from the
occurrence data — a useful baseline for the niche ellipsoid:

```{r ellipsoid}
pars_sample <- get_ellipsoid_pars(env_occ)
pars_sample$mu
pars_sample$s_mat
```

The fitted models refine these estimates via maximum likelihood. The
optimized parameter vector is stored in `fit$best$theta` and used
internally by `predict()` when projecting suitability onto rasters.

## 7 Project habitat suitability

The `predict()` method projects the fitted niche onto environmental
rasters, producing a geographic suitability map with values in $(0, 1]$.

```{r predict}
suit_po  <- predict(fit_po,  env_rasters)
suit_ipw <- predict(fit_ipw, env_rasters)
```

### Side-by-side comparison

```{r suitability-maps, fig.width = 10, fig.height = 4}
op <- par(mfrow = c(1, 2), mar = c(3, 3, 3, 5))
plot(suit_po,
     main = "Presence-only suitability",
     col = hcl.colors(50, "YlOrRd", rev = TRUE))
points(example_occ_df$lon, example_occ_df$lat, pch = 20, cex = 0.45)
plot(suit_ipw,
     main = "Inverse-weighted suitability",
     col = hcl.colors(50, "YlOrRd", rev = TRUE))
points(example_occ_df$lon, example_occ_df$lat, pch = 20, cex = 0.45)
par(op)
```

The inverse-weighted model typically produces a tighter niche estimate
because it down-weights regions of environmental space that are
over-represented in the background.

## 8 Iso-suitability contours in environmental space

```{r contours, fig.width = 7, fig.height = 5}
library(ggplot2)

ggplot() +
  geom_nicher_background(env_m, size = 0.35, alpha = 0.18) +
  geom_nicher_occ(env_occ, size = 1.2, alpha = 0.85) +
  geom_nicher_isosuitability(
    fit_po,
    level = c(0.75, 0.5, 0.25),
    colour = "#2f6fb2",
    linewidth = 0.7
  ) +
  geom_nicher_isosuitability(
    fit_ipw,
    level = c(0.75, 0.5, 0.25),
    colour = "#d95f02",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  labs(
    title = "Presence-only vs inverse-weighted niches",
    subtitle = "Blue = presence-only; orange dashed = inverse-weighted",
    x = names(env_occ)[1],
    y = names(env_occ)[2]
  ) +
  theme_minimal()
```

## References

Jimenez, L., Soberon, J., Christen, J. A., & Soto, D. (2022).
Dealing with critical uncertainty in ecological niche models.
*Ecological Modelling*, 468, 109823.
<https://doi.org/10.1016/j.ecolmodel.2021.109823>

Jimenez, L. & Soberon, J. (2022).
Weighted likelihood estimation of the fundamental niche.
*Ecological Modelling*, 470, 110009.
<https://doi.org/10.1016/j.ecolmodel.2022.110009>
