---
title: "Getting Started with nicher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with nicher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(nicher)
```

## Introduction

The **nicher** package estimates ecological niche models using ellipsoidal
distributions under the *M hypothesis* (Soberón & Peterson 2005). The niche is
represented as a multivariate normal distribution parameterised in the
unconstrained *math scale* (log-standard-deviations + C-vine partial
correlations), which allows unconstrained numerical optimization.

Seven likelihood formulations are supported:

| Model | Description | Reference |
|---|---|---|
| `"presence_only"` | Multivariate normal, no background | — |
| `"kde_bias_corrected"` | Legacy KDE-bias-corrected weighted normal | Jiménez & Soberón (2022) |
| `"weighted"` | Paper-faithful weighted normal with ridge regularization | Jiménez & Soberón (2022) |
| `"skew_normal"` | Presence-only multivariate skew-normal | Azzalini & Capitanio (1999) |
| `"skew_normal_weighted"` | Weighted multivariate skew-normal | Azzalini & Capitanio (1999); Jiménez & Soberón (2022) |
| `"skew_t"` | Presence-only non-central skew-t | Branco & Dey (2001) |
| `"skew_t_weighted"` | Weighted non-central skew-t | Branco & Dey (2001); Jiménez & Soberón (2022) |

---

## Built-in datasets

The package ships four environmental data frames for the hummingbird
*Abeillia abeillei* (Jiménez et al. 2022):

```{r data-overview}
# 2-D datasets (bio1 = temperature, bio12 = precipitation)
dim(example_env_occ_2d)   # presence points
dim(example_env_m_2d)     # M-hypothesis background sample

# 3-D datasets (adds bio19 = precipitation of coldest quarter)
dim(example_env_occ_3d)
dim(example_env_m_3d)

head(example_env_occ_2d)
```

---

## Starting values

`start_theta()` generates a single starting parameter vector from the sample
moments of the presence data. `start_theta_multiple()` generates a space-filling
design of multiple starts.

```{r start-theta}
theta0_2d <- start_theta(example_env_occ_2d)
length(theta0_2d)   # 2*p + p*(p-1)/2 = 2*2 + 1 = 5

theta0_3d <- start_theta(example_env_occ_3d)
length(theta0_3d)   # 2*3 + 3 = 9
```

---

## Baseline log-likelihood (`loglik_niche`)

`loglik_niche()` accepts a centroid vector `mu` and covariance matrix `s_mat`
directly. This is the pure-R reference implementation.

```{r loglik-niche}
ll <- loglik_niche(
  mu      = example_mu_vec,
  s_mat   = example_s_mat,
  env_occ = example_env_occ_2d,
  env_m   = example_env_m_2d,
  neg     = FALSE         # return the log-likelihood (not negative)
)
print(ll)
```

---

## Math-scale log-likelihood (`loglik_niche_math_cpp`)

`loglik_niche_math_cpp()` accepts the unconstrained parameter vector `theta`
(no matrix construction required) and delegates to a compiled C++ kernel for
efficiency.

```{r loglik-math-cpp}
ll_cpp <- loglik_niche_math_cpp(
  theta   = theta0_2d,
  env_occ = example_env_occ_2d,
  env_m   = example_env_m_2d,
  eta     = 1,
  neg     = FALSE
)
print(ll_cpp)
```

---

## Optimizing a weighted niche model

`optimize_niche()` runs multi-start optimization. The `breadth` parameter
controls the quantile range of starting points
(default `0.1` → quantiles `c(0.1, 0.5, 0.9)`).

```{r optimize-weighted}
set.seed(42)
res_w <- optimize_niche(
  env_occ         = example_env_occ_2d,
  env_m           = example_env_m_2d,
  num_starts      = 5L,
  breadth         = 0.1,
  likelihood      = "weighted",
  eta             = 1,
  m_subsample     = 500L,
  m_kde_subsample = 1000L,
  seed            = 1L
)

# Print the nicher object
print(res_w)

# Assess acceptance criteria
diag_w <- assess(res_w)
cat("Flag:", diag_w$flag, "\n")
cat("Recommendation:", diag_w$recommendation, "\n")
```

The fitted object stores the optimizer's best parameter vector in
`res_w$best$theta`. Its first entries are the fitted niche centroid (`mu`) on
the original environmental scale; the next entries are `log(sigma)`, followed
by the C-vine correlation parameters used internally.

```{r ellipsoid-pars}
theta_best <- res_w$best$theta
p <- length(res_w$var_names)
mu_hat <- theta_best[seq_len(p)]
print(mu_hat)
```

---

## Projecting habitat suitability

Use `predict()` with a `terra` raster stack whose layer names match the
environmental variables used to fit the model. The prediction is the
standardized Gaussian suitability surface:

\[
S(x) = \exp\left(-\frac{1}{2}(x - \mu)^\top \Sigma^{-1}(x - \mu)\right),
\]

so values are in `(0, 1]` and the fitted centroid has suitability 1.

```{r predict-suitability}
env_m_matrix <- as.matrix(example_env_m_2d)

template <- terra::rast(
  nrows = 40,
  ncols = 40,
  xmin  = min(env_m_matrix[, 1L]),
  xmax  = max(env_m_matrix[, 1L]),
  ymin  = min(env_m_matrix[, 2L]),
  ymax  = max(env_m_matrix[, 2L])
)

xy <- terra::crds(template, df = TRUE)
env_rast <- c(template, template)
terra::values(env_rast[[1L]]) <- xy[, 1L]
terra::values(env_rast[[2L]]) <- xy[, 2L]
names(env_rast) <- res_w$var_names

suitability <- predict(res_w, env_rast)
suitability
terra::global(suitability, c("min", "max"), na.rm = TRUE)
```

---

## Presence-only model

When no background data are available, use `likelihood = "presence_only"`.

```{r optimize-presence-only}
set.seed(42)
res_po <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  num_starts = 5L,
  breadth    = 0.1,
  likelihood = "presence_only",
  eta        = 1
)

print(res_po)
assess(res_po)$flag
```

---

## Non-symmetric models

Use the skew-normal or skew-t families when the niche response is
asymmetric rather than symmetric around the centroid. Skew-normal models
append one skew parameter `alpha` per environmental variable; skew-t
models add one more `log_r` degrees-of-freedom parameter for heavier
tails.

```{r optimize-skew-models}
set.seed(42)

res_sn <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  num_starts = 5L,
  breadth    = 0.1,
  likelihood = "skew_normal",
  eta        = 1
)

res_skt <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  num_starts = 5L,
  breadth    = 0.1,
  likelihood = "skew_t",
  eta        = 1
)

compare_nicher(
  presence_only = res_po,
  skew_normal = res_sn,
  skew_t = res_skt
)
```

For background-corrected non-symmetric models, use
`"skew_normal_weighted"` or `"skew_t_weighted"` with `env_m`.

```{r optimize-weighted-skew-models}
set.seed(42)

res_snw <- optimize_niche(
  env_occ         = example_env_occ_2d,
  env_m           = example_env_m_2d,
  num_starts      = 5L,
  breadth         = 0.1,
  likelihood      = "skew_normal_weighted",
  m_subsample     = 500L,
  m_kde_subsample = 1000L,
  seed            = 1L
)

res_sktw <- optimize_niche(
  env_occ         = example_env_occ_2d,
  env_m           = example_env_m_2d,
  num_starts      = 5L,
  breadth         = 0.1,
  likelihood      = "skew_t_weighted",
  m_subsample     = 500L,
  m_kde_subsample = 1000L,
  seed            = 1L
)

compare_nicher(
  weighted = res_w,
  skew_normal_weighted = res_snw,
  skew_t_weighted = res_sktw
)
```

See `vignette("non_symmetric_models", package = "nicher")` for a focused
walkthrough of skew-normal and skew-t parameter layouts.

---

## Low-level wrappers

For maximum performance with a fixed starting vector, use the model-specific
wrappers. `niche_presence_only()` handles presence-only fits, while
`niche_kde_bias_corrected()` exposes the legacy KDE-bias-corrected weighted
objective directly.

```{r low-level-wrappers}
occ3    <- as.matrix(example_env_occ_3d)
M3      <- as.matrix(example_env_m_3d)
theta3  <- start_theta(example_env_occ_3d)

res_po3 <- niche_presence_only(occ = occ3, start = theta3)
print(res_po3$value)
```

---

## References

Jiménez, L., & Soberón, J. (2019). Leaving the area under the ROC curve aside:
An alternative assessment of the performance of species distribution models.
*Ecological Modelling*, 402, 41–51.

Jiménez, L., Soberón, J., Christen, J. A., & Soto, D. (2022).
On the problem of modeling a fundamental niche from occurrence data.
*Ecological Modelling*, 397, 109823.
<https://doi.org/10.1016/j.ecolmodel.2021.109823>
