---
title: "Introduction to nicher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to nicher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `nicher` package estimates ecological niche models using ellipsoidal
geometry under an M hypothesis (Jimenez et al. 2022).
It fits seven likelihood families via multi-start optimisation with Sobol
sequences, backed by a compiled C++/Eigen engine for speed.

## Example data

The package ships with example datasets for *Vicugna vicugna*:

```{r data}
library(nicher)

str(example_env_occ_2d)
str(example_env_m_2d)
```

## Fitting a niche model

The main entry point is `optimize_niche()`. Here we fit a presence-only
model with a small number of starts for speed:
```{r fit-po}
set.seed(42)
fit_po <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  num_starts = 5L,
  breadth    = 0.1,
  likelihood = "presence_only"
)
print(fit_po)
```

## Convergence diagnostics

The `assess()` generic checks whether multiple starts converged to the
same optimum:
```{r assess}
diag <- assess(fit_po)
diag$flag
diag$recommendation
```

## Extracting ellipsoid parameters

Recover the estimated mean vector and covariance matrix on the natural
scale:
```{r ellipsoid}
pars <- get_ellipsoid_pars(example_env_occ_2d)
pars$mu
pars$s_mat
```

## Weighted model

For models that correct for sampling bias in the accessible area (M),
supply `env_m`:
```{r fit-w}
set.seed(42)
fit_w <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = example_env_m_2d,
  num_starts = 5L,
  breadth    = 0.1,
  likelihood = "weighted"
)
print(fit_w)
```

## Model comparison with AIC/BIC

Compare competing models using information criteria:
```{r compare, warning = FALSE}
cmp <- compare_nicher(
  presence_only = fit_po,
  weighted      = fit_w
)
print(cmp[, c("model", "likelihood", "AIC", "dAIC", "BIC", "dBIC")])
```

## References

Jimenez, L., Soberon, J., Christen, J. A., & Soto, D. (2022).
Dealing with critical uncertainty in ecological niche models.
*Ecological Modelling*, 468, 109823.
<https://doi.org/10.1016/j.ecolmodel.2021.109823>
