---
title: "Non-symmetric niche models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-symmetric niche models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(nicher)
```

## Why non-symmetric models?

The original nicher likelihoods represent a niche as a symmetric
multivariate Gaussian ellipsoid. That is a useful baseline, but
ecological responses are not always symmetric: suitability can decline
slowly on one side of an optimum and sharply on the other. nicher
therefore includes two non-symmetric families:

| Family | Background correction | Extra parameters |
|---|---:|---|
| `"skew_normal"` | No | `alpha_1, ..., alpha_p` |
| `"skew_normal_weighted"` | Yes | `alpha_1, ..., alpha_p` |
| `"skew_t"` | No | `alpha_1, ..., alpha_p`, `log_r` |
| `"skew_t_weighted"` | Yes | `alpha_1, ..., alpha_p`, `log_r` |

`alpha` controls direction and strength of skew. `alpha = 0` recovers
the symmetric Gaussian sub-model. The skew-t families add the degrees of
freedom parameter `r`, stored as `log_r` on the unconstrained math scale;
smaller `r` gives heavier tails and large `r` approaches the
skew-normal.

## Example data

This vignette uses small subsets of the bundled hummingbird data so it
knits quickly. For analyses, fit the full data and increase
`num_starts` and `control$maxeval`.

```{r data}
env_occ <- example_env_occ_2d[seq_len(40), ]
env_m   <- example_env_m_2d[seq_len(500), ]

dim(env_occ)
dim(env_m)
head(env_occ)
```

## Starting values and parameter layout

The Gaussian model has `mu`, `log_sigma`, and C-vine correlation
parameters. The skew-normal appends one `alpha` per variable. The skew-t
then appends `log_r`.

```{r starts}
theta_gaussian <- start_theta(env_occ)
theta_skew     <- start_theta(env_occ, skew = TRUE)
theta_skew_t   <- start_theta(env_occ, skew_t = TRUE)

c(
  gaussian = length(theta_gaussian),
  skew_normal = length(theta_skew),
  skew_t = length(theta_skew_t)
)

names(start_theta_multiple(env_occ, num_starts = 2L, skew_t = TRUE))
```

## Fit presence-only skew models

Use `"skew_normal"` or `"skew_t"` when you want a non-symmetric model
from occurrence data only.

```{r fit-presence-skew}
doc_control <- list(maxeval = 150L)

fit_po <- optimize_niche(
  env_occ = env_occ,
  env_m = NULL,
  num_starts = 3L,
  likelihood = "presence_only",
  seed = 10L,
  control = doc_control,
  verbose = FALSE
)

fit_sn <- optimize_niche(
  env_occ = env_occ,
  env_m = NULL,
  num_starts = 3L,
  likelihood = "skew_normal",
  seed = 10L,
  control = doc_control,
  verbose = FALSE
)

fit_st <- optimize_niche(
  env_occ = env_occ,
  env_m = NULL,
  num_starts = 3L,
  likelihood = "skew_t",
  seed = 10L,
  control = doc_control,
  verbose = FALSE
)

compare_nicher(
  gaussian = fit_po,
  skew_normal = fit_sn,
  skew_t = fit_st
)
```

## Fit weighted skew models

Use `"skew_normal_weighted"` or `"skew_t_weighted"` when you want
non-symmetric responses with the same *M*-background correction used by
the weighted Gaussian family.

```{r fit-weighted-skew}
fit_w <- optimize_niche(
  env_occ = env_occ,
  env_m = env_m,
  num_starts = 3L,
  likelihood = "weighted",
  seed = 10L,
  m_subsample = nrow(env_m),
  m_kde_subsample = nrow(env_m),
  control = doc_control,
  verbose = FALSE
)

fit_snw <- optimize_niche(
  env_occ = env_occ,
  env_m = env_m,
  num_starts = 3L,
  likelihood = "skew_normal_weighted",
  seed = 10L,
  m_subsample = nrow(env_m),
  m_kde_subsample = nrow(env_m),
  control = doc_control,
  verbose = FALSE
)

fit_stw <- optimize_niche(
  env_occ = env_occ,
  env_m = env_m,
  num_starts = 3L,
  likelihood = "skew_t_weighted",
  seed = 10L,
  m_subsample = nrow(env_m),
  m_kde_subsample = nrow(env_m),
  control = doc_control,
  verbose = FALSE
)

compare_nicher(
  weighted = fit_w,
  skew_normal_weighted = fit_snw,
  skew_t_weighted = fit_stw
)
```

## Extract skew parameters

The helper below reads the `alpha` and `r` blocks from a fitted model.

```{r extract-skew}
extract_skew <- function(fit) {
  p <- length(fit$var_names)
  n_cor <- p * (p - 1L) / 2L
  theta <- fit$best$theta
  alpha_start <- 2L * p + n_cor + 1L

  alpha <- theta[alpha_start + seq_len(p) - 1L]
  names(alpha) <- paste0("alpha_", fit$var_names)

  r <- if (grepl("skew_t", fit$likelihood)) {
    exp(theta[alpha_start + p])
  } else {
    NA_real_
  }

  c(alpha, r = r)
}

rbind(
  skew_normal = extract_skew(fit_sn),
  skew_t = extract_skew(fit_st),
  skew_normal_weighted = extract_skew(fit_snw),
  skew_t_weighted = extract_skew(fit_stw)
)
```

## Prediction

`predict()` uses the fitted centroid and covariance to return the same
standardized ellipsoidal suitability surface for all families. The skew
parameters affect fitting, model comparison, and selected `theta`, but
the current projection is still an ellipsoid in environmental space.

```{r predict-skew}
env_m_matrix <- as.matrix(env_m)

template <- terra::rast(
  nrows = 40,
  ncols = 40,
  xmin = min(env_m_matrix[, 1L]),
  xmax = max(env_m_matrix[, 1L]),
  ymin = min(env_m_matrix[, 2L]),
  ymax = max(env_m_matrix[, 2L])
)

xy <- terra::crds(template, df = TRUE)
env_rast <- c(template, template)
terra::values(env_rast[[1L]]) <- xy[, 1L]
terra::values(env_rast[[2L]]) <- xy[, 2L]
names(env_rast) <- fit_snw$var_names

skew_suitability <- predict(fit_snw, env_rast)
terra::global(skew_suitability, c("min", "max"), na.rm = TRUE)
```

## Practical guidance

* Start with `"presence_only"` and `"weighted"` as symmetric baselines.
* Add `"skew_normal"` or `"skew_normal_weighted"` when occurrence data
  suggest asymmetric environmental responses.
* Add `"skew_t"` or `"skew_t_weighted"` when a few observations may be
  from heavier-tailed responses.
* Keep `prior_alpha_lambda > 0` unless you deliberately need the pure
  maximum-likelihood skew fit; mild shrinkage stabilizes the direct
  skew parameterization.
* Use `cv_nicher()` to tune penalties before final model comparison.

## References

Azzalini, A., & Capitanio, A. (1999). Statistical applications of the
multivariate skew normal distribution. *Journal of the Royal Statistical
Society: Series B*, 61(3), 579-602.

Branco, M. D., & Dey, D. K. (2001). A general class of multivariate
skew-elliptical distributions. *Journal of Multivariate Analysis*,
79(1), 99-113.

Jiménez, L., Soberón, J., Christen, J. A., & Soto, D. (2022).
On the problem of modeling a fundamental niche from occurrence data.
*Ecological Modelling*, 397, 109823.
<https://doi.org/10.1016/j.ecolmodel.2021.109823>
