---
title: "Non-Symmetric Models: Skew-Normal, Skew-t, and NCST Families"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-Symmetric Models: Skew-Normal, Skew-t, and NCST Families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Motivation

Symmetric Gaussian models assume that a species responds equally to
deviations above and below its environmental optimum. Many species
violate this assumption: heat tolerance may extend further than cold
tolerance, or precipitation responses may be right-skewed.

`nicher` provides three non-symmetric likelihood families that relax the
symmetry assumption while retaining the ellipsoidal geometry:

| Family | Parameters | Tail weight | Background correction |
|--------|-----------|------------|----------------------|
| `skew_normal` | $3p + p(p-1)/2$ | Gaussian | No |
| `skew_t` | $3p + p(p-1)/2 + 1$ | Heavy | No |
| `ncst` | $3p + p(p-1)/2 + 1$ | Heavy | No |
| `skew_normal_weighted` | $3p + p(p-1)/2$ | Gaussian | Yes (KDE) |
| `skew_t_weighted` | $3p + p(p-1)/2 + 1$ | Heavy | Yes (KDE) |
| `ncst_weighted` | $3p + p(p-1)/2 + 1$ | Heavy | Yes (KDE) |

## 1 The skew-normal family

The skew-normal distribution [@azzalini1999statistical] adds a vector
$\alpha \in \mathbb{R}^p$ of skewness parameters to the multivariate
normal. When $\alpha = 0$, the model reduces to the symmetric Gaussian.

### Fitting

```{r fit-sn, results = "hold"}
library(nicher)
set.seed(42)

fit_sn <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  likelihood = "skew_normal",
  num_starts = 5L,
  breadth    = 0.1,
  control    = list(maxeval = 1000L),
  verbose    = FALSE
)
fit_sn
```

The estimated skewness parameters indicate the direction and magnitude
of asymmetry:

```{r sn-alpha}
theta <- fit_sn$best$theta
p <- ncol(example_env_occ_2d)
n_v <- p * (p - 1L) / 2L
alpha <- theta[(2L * p + n_v + 1L):(3L * p + n_v)]
names(alpha) <- colnames(example_env_occ_2d)
alpha
```

## 2 The skew-t family

The skew-t model [@branco2001general] extends the skew-normal by
replacing Gaussian tails with Student-$t$ tails controlled by a
degrees-of-freedom parameter $r > 0$:

$$
T = \mu + X\sqrt{r/Y}, \qquad X \sim \text{SN}_p(0, \Sigma, \alpha),
\quad Y \sim \chi^2_r.
$$

As $r \to \infty$, the skew-t converges to the skew-normal.

### Fitting

```{r fit-st, results = "hold"}
set.seed(42)

fit_st <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  likelihood = "skew_t",
  num_starts = 5L,
  breadth    = 0.1,
  control    = list(maxeval = 1000L),
  verbose    = FALSE
)
fit_st
```

The estimated degrees of freedom:

```{r st-df}
log_r <- fit_st$best$theta[length(fit_st$best$theta)]
cat("Estimated r =", round(exp(log_r), 2), "\n")
```

## 3 The NCST family

The Non-Central Skew-$t$ (NCST) places the location parameter $\xi$
*inside* the skew-normal component before chi-squared scaling:

$$
T = \frac{X}{\sqrt{Y/r}}, \qquad X \sim \text{SN}_p(\xi, \Omega, \alpha),
\quad Y \sim \chi^2_r.
$$

This allows degrees of freedom to interact with the location-scale
structure, providing extra flexibility compared to the standard skew-t.

### Fitting

```{r fit-ncst, results = "hold"}
set.seed(42)

fit_ncst <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = NULL,
  likelihood = "ncst",
  num_starts = 5L,
  breadth    = 0.1,
  control    = list(maxeval = 1000L),
  verbose    = FALSE
)
fit_ncst
```

## 4 Model comparison

Compare all non-symmetric families using AIC and BIC:

```{r compare, warning = FALSE}
cmp <- compare_nicher(
  skew_normal = fit_sn,
  skew_t      = fit_st,
  ncst        = fit_ncst
)
print(cmp[, c("model", "likelihood", "AIC", "dAIC", "BIC", "dBIC")])
```

## 5 Weighted variants

Each non-symmetric family has a weighted counterpart that corrects for
sampling bias using a KDE of the background environment. Supply `env_m`
and use the `_weighted` suffix:

```{r fit-weighted, results = "hold"}
set.seed(42)

fit_snw <- optimize_niche(
  env_occ    = example_env_occ_2d,
  env_m      = example_env_m_2d,
  likelihood = "skew_normal_weighted",
  num_starts = 5L,
  breadth    = 0.1,
  control    = list(maxeval = 500L),
  verbose    = FALSE
)
fit_snw
```

## 6 When to use non-symmetric models

- **Skew-normal**: When the species has clearly asymmetric environmental
  tolerances but Gaussian-like tails.
- **Skew-t**: When the species tolerates extreme environmental values
  more than a Gaussian would predict (heavy tails).
- **NCST**: When additional flexibility in how tail heaviness interacts
  with location is needed. Compare with skew-t via AIC/BIC.
- **Weighted variants**: When sampling bias in the background
  environment must be corrected simultaneously with asymmetry.

Use `compare_nicher()` or `cv_nicher()` to select among families
for a given dataset.

## References
